Secure Composition of Robust and Optimising Compilers

### 2023-07-10 popl/main.tex
make MK visible

In summary, this paper makes the following contributions:
\begin{itemize}
  \item %
        %\Cref{sec:compprop} presents a formal framework to reason about compositions of security properties and demonstrate its usage with concrete examples:
        First, this paper studies the formalisation and the composition of security properties (\Cref{sec:compprop}) by focussing on those properties that are of interest for real-world compiler writers (as identified by the plethora of work enforcing such properties individually~\cite{akritidis2009baggy,nagarakatte2009soft,nagarakatte2010cets,dhumbumroong2020boundwarden,jung2021pico,nam2019framer,shankaranarayana2023tailcheck,younan2010paricheck,zhou2023fatptrs,bond2017vale,almeida2017jasmin,kuepper2023cryptopt,cauligi2019fact}).
        % 
        Starting from ways to formalise and enforce those properties individually, this paper shows how to compose their formalisation. % and their composition. %% MK: compose their composition??
        % 
        This paper then showcases the benefits of this composition by composing \gls{tms} and \gls{sms} into \gls{ms}, and ultimately adding \gls{cct} to the composition, yielding the the conjunction of \gls{ms} and \gls{cct}.
        % 
        The resulting security property is the golden standard of safety properties that secure cryptographic code should have~\cite{lemay2021ccc}.
%        \MPin{
%        	conclude that did this prop in a single pass, we show how to do it compositionally
%        }

@@

Property satisfaction is defined on whole programs, i.e., programs without missing definitions.
Thus, there is only a passive attacker model that observes the execution and, e.g., retreives secrets from that.
To consider a stronger model similarly to what existing work does~\cite{abate2019jour,abate2021extacc,patrignani2021rsc}, the concept of satisfaction is extended with {\em robustness}.
Robust satisfaction considers partial programs $\progvar$, i.e., programs with free variables, which need to be defined prior to execution.
Programs with free variables can get stuck during execution.
To remedy this, {\em linking} takes two partial programs $\progvar[_{1}],\progvar[_{2}]$ and produces a whole program $\wholeprogvar$, i.e., $\link{\progvar[_{1}]}{\progvar[_{2}]}=\wholeprogvar$.
As typically done in not only secure compilation~\cite{abate2019jour,devriese2018parametricity}, %
% \MP{others do this too, parametricity results e.g.}
this paper assumes that whole programs are the result of linking partial programs referred to as {\em context} ($\ctx$) and {\em component} ($\comp$).
The context is an arbitrary program and thus has the role of an {\em attacker} that can interact with the component by means of whatever features the programming language has.
%In this work, the semantics of the programming language is expected to differentiate between the component and the context.
With this, \Thmref{def:propsat} can be extended as follows: for \bul{components $\progvar$ to robustly satisfy a property $\pi$}, take an \iul{attacker context $\contextvar$ and link it with $\progvar$}, \oul{the result of which must satisfy $\pi$}.
%\MPin{
% the behaviour of the result?
%%%%%%%    MK: def:propsat defines satisfaction in terms of programs, no need to talk about behavior here
%}

@@

A {\em compiler} ($\cc{\src{L}}{\trg{L}}$) translates syntactic descriptions of programs from a {\em source} ($\src{L}$) into a {\em target} ($\trg{L}$) programming language.
%For the syntactic description of programs written in $\src{L}$, this paper writes $\src{\progvar}$, while for ones written in $\trg{L}$, this paper writes $\trg{\progvar}$.%\MP{not a fan of this sentence}
%This paper writes $\src{\progvar}$ and $\trg{\progvar}$ for programs written in $\src{L}$ and $\trg{L}$, respectively.
%      MK: I don't think we need to mention this at all
This translation is considered {\em correct} if it is semantics-preserving.
That is, for a whole program $\src{\wholeprogvar}$, the compiler should relate the $\src{L}$ semantics of $\src{\wholeprogvar}$ with the semantics of $\trg{T}$ in such a way that they are ,,compatible''.
Unfortunately, correct compilers may be insecure compilers~\cite{patrignani2019survey}, and programs translated by insecure compilers can violate security properties that the programmer assumes to hold.
To define when a compiler is secure, this paper uses the robust compilation framework~\cite{abate2019jour}, which we summarize in the following definition.

@@

\subsubsection{Securing Optimisations}
%\MPin{note that i'm always using british spelling. optimisation VS optimization. etc. }
% MK: I'll go over the doc with the antidote spellchecker as soon as it is reasonably stable.

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
%\MPin{
%	do not say the negation of something before the non-negation.
%	say what things are first, and what they are not later (if at all)
%}  MK: that for sure is less confusing... Good suggestion!
Both examples can be seen in practice~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion}, it is infeasible to list them all\MP{survey to cite? this is a bit informal}. % MK: hard to find a survey on different compiler passes... I can't even find one just for memsafety
Consider the compilation passes of LLVM~\cite{lattner2004llvm}.
In LLVM, compiler passes are handled in the \texttt{PassManager} class.
By means of manipulating the internal state of the \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
That way, an order of optimisations can be found by hand that yields the best results.
Take \gls{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls{dce}, which removes dead code by rewriting conditional branches.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

Above is a fraction of the shared syntax of all the programming languages of this paper.
\MP{why a fraction? what exprs are missing}
%% MK: Things that are missing here are: Evaluation contexts, states Ω and transitively other definitions affected by this
The syntax and semantics are mostly standard and therefore omitted.
However, when allocating arrays with $\lnew{\expr[_{1}]}{\expr[_{2}]}$, the expression $\expr[_{1}]$ determines the initial value of the cells and $\expr[_{2}]$ the overall size of an array.

\MPin{
	below is semantics info. should go with their symbols and the semantics rules.
	and a better description of the semantics symbols.
% MK: this moved to appendix! should we bring it back?
}
The data associated to pointers resides in the dynamic state $\Omega$ that contains sandboxed heaps, one for the context and the other for the component.
Because of the sandboxing, locations pointing into heaps must be associated with a control-tag $\sandboxtag$ (either $\ctx$ or $\comp$) to know the heap they point into.
Other metadata in $\Omega$ entails a list of functions with information on whether they belong to the context or the component.
That way, when calling a function, the semantics can take care of emitting appropriate events that signal a context switch.
For the languages considered in this paper, functions are second-class.
So, the only values are pairs, pointers, and natural numbers.
Another non-standard part of the languages is the concept of the poison-tag.
The poison-tag marks an array as freed ($\poisoned$) without removing the information associated to the freed pointer from $\Omega$.
This measure prevents the semantics from getting stuck when an already freed pointer is freed again.
Because of this, it is possible that a program like $\llet{x}{\lnew{0}{5}}{\ldelete{x};\ldelete{x}}$ yields a trace $\ev{Alloc\ \loc\ \sandboxtag\ 5}\cdot\ev{Dealloc\ \loc\ \sandboxtag}\cdot\ev{Dealloc\ \loc\ \sandboxtag}$ when executing.

>>>>

In summary, this paper makes the following contributions:
\begin{itemize}
  \item %
        %\Cref{sec:compprop} presents a formal framework to reason about compositions of security properties and demonstrate its usage with concrete examples:
        First, this paper studies the formalisation and the composition of security properties (\Cref{sec:compprop}) by focussing on those properties that are of interest for real-world compiler writers (as identified by the plethora of work enforcing such properties individually~\cite{akritidis2009baggy,nagarakatte2009soft,nagarakatte2010cets,dhumbumroong2020boundwarden,jung2021pico,nam2019framer,shankaranarayana2023tailcheck,younan2010paricheck,zhou2023fatptrs,bond2017vale,almeida2017jasmin,kuepper2023cryptopt,cauligi2019fact}).
        % 
        Starting from ways to formalise and enforce those properties individually, this paper shows how to compose their formalisation.
        % 
        This paper then showcases the benefits of this composition by composing \gls{tms} and \gls{sms} into \gls{ms}, and ultimately adding \gls{cct} to the composition, yielding the the conjunction of \gls{ms} and \gls{cct}.
        % 
        The resulting security property is the golden standard of safety properties that secure cryptographic code should have~\cite{lemay2021ccc}.
%        \MPin{
%        	conclude that did this prop in a single pass, we show how to do it compositionally
%        }

@@

Property satisfaction is defined on whole programs, i.e., programs without missing definitions.
Thus, there is only a passive attacker model that observes the execution and, e.g., retreives secrets from that.
To consider a stronger model similarly to what existing work does~\cite{abate2019jour,abate2021extacc,patrignani2021rsc}, the concept of satisfaction is extended with {\em robustness}.
Robust satisfaction considers partial programs $\progvar$, i.e., programs with free variables, which need to be defined prior to execution.
Programs with free variables can get stuck during execution.
To remedy this, {\em linking} takes two partial programs $\progvar[_{1}],\progvar[_{2}]$ and produces a whole program $\wholeprogvar$, i.e., $\link{\progvar[_{1}]}{\progvar[_{2}]}=\wholeprogvar$.
As typically done in not only secure compilation~\cite{abate2019jour,devriese2018parametricity}, %
% \MP{others do this too, parametricity results e.g.}
this paper assumes that whole programs are the result of linking partial programs referred to as {\em context} ($\ctx$) and {\em component} ($\comp$).
The context is an arbitrary program and thus has the role of an {\em attacker} that can interact with the component by means of whatever features the programming language has.
%In this work, the semantics of the programming language is expected to differentiate between the component and the context.
With this, \Thmref{def:propsat} can be extended as follows: for \bul{components $\progvar$ to robustly satisfy a property $\pi$}, take an \iul{attacker context $\contextvar$ and link it with $\progvar$}, \oul{the result of which must satisfy $\pi$}.

@@

A {\em compiler} ($\cc{\src{L}}{\trg{L}}$) translates syntactic descriptions of programs from a {\em source} ($\src{L}$) into a {\em target} ($\trg{L}$) programming language.
This translation is considered {\em correct} if it is semantics-preserving.
That is, for a whole program $\src{\wholeprogvar}$, the compiler should relate the $\src{L}$ semantics of $\src{\wholeprogvar}$ with the semantics of $\trg{T}$ in such a way that they are ,,compatible''.
Unfortunately, correct compilers may be insecure compilers~\cite{patrignani2019survey}, and programs translated by insecure compilers can violate security properties that the programmer assumes to hold.
To define when a compiler is secure, this paper uses the robust compilation framework~\cite{abate2019jour}, which we summarize in the following definition.

@@

\subsubsection{Securing Optimisations}
Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion}, it is infeasible to list them all\MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
Consider the compilation passes of LLVM~\cite{lattner2004llvm}.
In LLVM, compiler passes are handled in the \texttt{PassManager} class.
By means of manipulating the internal state of the \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
That way, an order of optimisations can be found by hand that yields the best results.
Take \gls{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls{dce}, which removes dead code by rewriting conditional branches.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

Above is a fraction of the shared syntax of all the programming languages of this paper.
\MPin{why a fraction? what exprs are missing
MK: Things that are missing here are: Evaluation contexts, states Ω and transitively other definitions affected by this
}
The syntax and semantics are mostly standard and therefore omitted.
However, when allocating arrays with $\lnew{\expr[_{1}]}{\expr[_{2}]}$, the expression $\expr[_{1}]$ determines the initial value of the cells and $\expr[_{2}]$ the overall size of an array.

\MPin{
	below is semantics info. should go with their symbols and the semantics rules.
	and a better description of the semantics symbols.
  MK: this moved to appendix! should we bring it back? what exactly should be left out?
}
The data associated to pointers resides in the dynamic state $\Omega$ that contains sandboxed heaps, one for the context and the other for the component.
Because of the sandboxing, locations pointing into heaps must be associated with a control-tag $\sandboxtag$ (either $\ctx$ or $\comp$) to know the heap they point into.
Other metadata in $\Omega$ entails a list of functions with information on whether they belong to the context or the component.
That way, when calling a function, the semantics can take care of emitting appropriate events that signal a context switch.
For the languages considered in this paper, functions are second-class.
So, the only values are pairs, pointers, and natural numbers.
Another non-standard part of the languages is the concept of the poison-tag.
The poison-tag marks an array as freed ($\poisoned$) without removing the information associated to the freed pointer from $\Omega$.
This measure prevents the semantics from getting stuck when an already freed pointer is freed again.
Because of this, it is possible that a program like $\llet{x}{\lnew{0}{5}}{\ldelete{x};\ldelete{x}}$ yields a trace $\ev{Alloc\ \loc\ \sandboxtag\ 5}\cdot\ev{Dealloc\ \loc\ \sandboxtag}\cdot\ev{Dealloc\ \loc\ \sandboxtag}$ when executing.
